Probing the Milky Way's Dark Matter Halo with RR Lyraes in Tidal Streams I: The Sagittarius Leading Arm

### 2013-06-25 text/proposals/mdm_proposal_may2013.tex
move docs to text

\begin{document}
\title{Probing the Milky Way's Dark Matter Halo with RR Lyraes in Tidal Streams I: The Sagittarius Leading Arm}
\author{Adrian Price-Whelan\altaffilmark{1}, Kathryn Johnston\altaffilmark{1}, Allyson Sheffield\altaffilmark{1}}
\altaffiltext{1}{Department of Astronomy, Columbia University, Mail Code 5246, New York, NY 10027}
Galaxies, by mass, are mostly composed of \emph{Dark Matter}. Simulations of structure and galaxy formation -- both with
and without baryons -- lead us to believe that such \emph{halos} of
dark matter have universal density profiles \citep{nfw96} and are triaxial
in shape \citep{jing02}. Including baryons in such simulations tends to
soften the triaxiality and alters the radial density profile. Thus,
precise measurements of the shape, orientation, and radial profile of
dark matter halos provides information about both dark matter physics
and the baryonic processes that helped shape them. Tidal streams from
infalling satellite galaxies have been used to constrain properties of
the Milky Way's halo \citep[e.g.][]{law10}; typically found outwards of 20~kpc, the
orbits of associated stars are still largely under the influence of
the dark matter and thus serve as potentiometers \citep[e.g.][]{johnston99}. Using an updated
method that utilizes full phase-space (6D) kinematic information for a
small sample ($\sim$100s) of stars stripped from a known progenitor
(e.g., Sagittarius; Price-Whelan et al 2013, in prep), it is
possible to accurately constrain the axis ratios, total mass, and
orientation of the Milky Way's dark matter halo. This method relies on a technique for deriving
distances to individual RR Lyrae stars with 2\% accuracy \citep{},
along with anticipated GAIA proper motion errors of
$\sim10-20~\mu\mathrm{as}/\mathrm{yr}$ for such stars at $\sim$
20-30~kpc. This leaves one remaining velocity component -- the radial
velocity.  We plan to observe a
sample of RR Lyrae stars selected to be likley members of Sgr's debris system 
using catalogs of RRab stars from
time-domain, photometric surveys such as LINEAR \citep{}, PTF \citep{}
and CSS \citep{}. 
Our observations will both confirm membership of the debris system in order to construct a target list for Spitzer (proposals are due in August 2013) as well
as measure the missing velocity dimension.

@@

\begin{thebibliography}[99]
\bibitem[Jing 
\& Suto(2002)]{jing02} Jing, Y.~P., \& Suto, Y.\ 2002, \apj, 574, 538 
\bibitem[Navarro et al.(1996)]{nfw96} Navarro, J.~F., Frenk, 
C.~S., \& White, S.~D.~M.\ 1996, \apj, 462, 563 
\end{thebibliography}

>>>>

\begin{document}
\title{Probing the Milky Way's Dark Matter Halo with RR Lyraes in
  Tidal Streams I: The Sagittarius Leading Arm} \author{Adrian M.
  Price-Whelan\altaffilmark{1}, Kathryn Johnston\altaffilmark{1},
  Allyson Sheffield\altaffilmark{1}} \altaffiltext{1}{Department of
  Astronomy, Columbia University, Mail Code 5246, New York, NY 10027}

Galaxies, by mass, are mostly composed of \emph{Dark
  Matter}. Simulations of structure and galaxy formation -- both with
and without baryons -- lead us to believe that such \emph{halos} of
dark matter have universal density profiles \citep{nfw96} and are
triaxial in shape \citep{jing02}. Including baryons in such
simulations tends to soften the triaxiality and alters the radial
density profile. Thus, precise measurements of the shape, orientation,
and radial profile of dark matter halos provides information about
both dark matter physics and the baryonic processes that helped shape
them. Tidal streams from infalling satellite galaxies have been used
to constrain properties of the Milky Way's halo \citep[e.g.][]{law10};
typically found outwards of 20~kpc, the orbits of associated stars are
still largely under the influence of the dark matter and thus serve as
potentiometers \citep[e.g.][]{johnston99}. Using an updated method
that utilizes full phase-space (6D) kinematic information for a small
sample ($\sim$100s) of stars stripped from a known progenitor (e.g.,
Sagittarius; Price-Whelan et al 2013, in prep), it is possible to
accurately constrain the axis ratios, total mass, and orientation of
the Milky Way's dark matter halo. This method relies on a technique
for deriving distances to individual RR Lyrae stars with 2\% accuracy
\citep{madore12}, along with anticipated GAIA proper motion errors of
$\sim10-20~\mu\mathrm{as}/\mathrm{yr}$ for such stars at $\sim$
20-30~kpc. This leaves one remaining velocity component -- the radial
velocity.  We plan to observe a sample of RR Lyrae stars selected to
be likley members of Sgr's debris system using catalogs of RRab stars
from time-domain, photometric surveys such as LINEAR \citep{linear}, PTF
\citep{ptf} and CSS \citep{catalina}.  Our observations will both confirm
membership of the debris system in order to construct a target list
for Spitzer (proposals are due in August 2013) as well as measure the
missing velocity dimension.

@@

\begin{thebibliography}{ }
\bibitem[Jing \& Suto(2002)]{jing02} Jing, Y.~P., \& Suto, Y.\ 2002, \apj, 574, 538
\bibitem[Navarro et al.(1996)]{nfw96} Navarro, J.~F., Frenk, C.~S., \& White, S.~D.~M.\ 1996, \apj, 462, 563
\bibitem[Law \& Majewski(2010)]{law10} Law, D.~R., \& Majewski, S.~R.\ 2010, \apj, 714, 229 
\bibitem[Johnston et al.(1999)]{johnston99} Johnston, K.~V., Zhao, H., Spergel, D.~N., \& Hernquist, L.\ 1999, \apjl, 512, L109 
\bibitem[Madore \& Freedman(2012)]{madore12} Madore, B.~F., \& Freedman, W.~L.\ 2012, \apj, 744, 132 
\bibitem[Stokes et al.(2000)]{linear} Stokes, G.~H., Evans, J.~B., Viggh, H.~E.~M., Shelly, F.~C., \& Pearce, E.~C.\ 2000, \icarus, 148, 21 
\bibitem[Law et al.(2009)]{ptf} Law, N.~M., Kulkarni, S.~R., Dekany, R.~G., et al.\ 2009, \pasp, 121, 1395 
\bibitem[Drake et al.(2013)]{catalina} Drake, A.~J., Catelan, M., Djorgovski, S.~G., et al.\ 2013, \apj, 763, 32 
\end{thebibliography}
